---
title: "Computational Sociology" 
subtitle: "Data structures"
author: Dr. Thomas Davidson
institute: Rutgers University
date: January 25, 2021
output:
    beamer_presentation:
      theme: "Szeged"
      colortheme: "beaver"
      fonttheme: "structurebold"
      toc: false
      incremental: false
header-includes:
  - \usepackage{multicol}
  - \usepackage{caption}
  - \captionsetup[figure]{font=scriptsize}
  - \captionsetup[figure]{labelformat=empty}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(dev = 'pdf')
library("knitr")
library("formatR")

opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
opts_chunk$set(tidy = FALSE)

knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```


# Plan
- Object-oriented programming
- Basic types
- Vectors
- Lists
- Matrices
- Data frames and tibbles
- A note on style

# Object-oriented programming
- A paradigm of computer programming
    - We create *objects* of different *classes* such as numbers, strings, and data frames
    - These objects have *attributes*, properties such as data
        - e.g. The numeric object we call **`A`** has an attribute called `value` equal to **`1`**
    - Objects are associated with *methods* that allow us to manipulate them
        - e.g. a numeric object might have a method called `add`, such that **`A + A`** will return **`2`**.

# Basic types
There are four basic types we will be using throughout the class. We use the `<-` operator to assign an object to a name.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
# Character (also known as called "strings")
name <- "Tom"
# Numeric ("float" in Python)
height <- 6.1
# Integer ("int" for short)
age <- 32L
# Logical
human <- TRUE
```
\tiny The other two are called `complex` and `raw`. See documentation: https://cran.r-project.org/doc/manuals/R-lang.html


# Basic types
There are a few useful commands for inspecting objects.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
print(name)
class(name)
typeof(name)
length(name)
attributes(name)
```

# Basic types
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
print(height)
class(height)
typeof(height)
length(height)
attributes(height)
```

# Basic types
We can also use the `==` expression to verify the content of an object. For numeric values we can also use some other expressions We will cover Boolean logic and truth statements more next lecture.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
name == "Tom"
age == 33L
age >= 30L # is greater than
age != 33L # is not
```

# Vectors
A vector is a collection of elements of the same class
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
# We can define an empty vector with N elements of a class
x <- logical(5)
print(x)
y <- numeric(5)
print(y)
z <- character(5)
print(z)
```

# Vectors
Let's take a closer look at numeric vectors.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v1 <- c(1,2,3,4,5)
v2 <- c(1,1,1,1,1)
class(v1) # check the class of this vectr
v1 + v2 # addition
v1 * v2 # multiplication
v1 - v2 # subtraction
sum(v1) # sum over v1
```
Note how the different methods return different outputs. The arithmetic operations return a new vector while `sum` returns a numeric value.

# Vectors
There are lots of commands for generating special types of numeric vectors. For example,
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
N <- 10

seq(N) # generates a sequence from 1 to N
rev(seq(N)) # reverses a vector
rnorm(N) # samples N times from a normal distribution
rbinom(N,1,0.5) # N observations of a single trial with a 0.5 probability of success
```

# Vectors
We can use the help `?` command to find information about each of these commands.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
?rnorm
```

# Vectors
We can use the `index` to access the specific elements of a vector. R uses square brackets for such indexing.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
x <- rnorm(N)
print(x)
print(x[1]) # R indexing starts at 1; Python and some others start at 0
x[1] <- 9 # We can also use indexing to modify elements
print(x[1])
```

# Vectors
The `head` and `tail` commands are useful when we're working with larger objects.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
x <- rnorm(10000)
length(x)
head(x)
tail(x)
head(x, n=20)
```

# Vectors
Vectors can also contain `null` elements to indicate missing values, represented by the `NA` string.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
x <- c(1,2,NULL)
x
```

# Lists
A list is an object that can contain different types of elements including basic types and vectors.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v1.list <- list(v1) # We can easily convert the vector v1 into a list.
v1.list[1] # The entire vector is considered the first element of the list
v1.list[[1]][1] # Double brackets then single to access a specific element
v1.list[1][1] # If first brackets are not double we just get the whole vector 
```
\tiny Note that indexing lists can be confusing so stick to vectors if possible.

# Lists
We can easily combine multiple vectors into a list.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v.list <- list(v1,v2) # We could store both vectors in a list
v.list[[1]][4] # We can use double brackets to get element 4 of list 1
```

# Lists
We can make indexing easier if we start with an empty list and then add elements using a named index.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v <- list() # initialize empty list
v$v1 <- v1 # the $ sign allows for named indexing
v$v2 <- v2
print(v)
```

# Lists
We can then use the `$` to index elements of the list.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v$v1
v$v1[4] # still need to use brackets to access a specific element
```
\tiny See https://cran.r-project.org/doc/manuals/R-lang.html#Indexing for more on indexing.


# Lists
A list could contain a mix of different types. These can be handy data structures but can also get complex very quickly.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
L <- list(name, numeric(5), TRUE, c("a","b","c"))
print(L)
```

# Matrices
A matrix is a two-dimensional data structure.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
M <- matrix(nrow=5,ncol=5) # Here there is no content so the matrix is empty
M
```

# Matrices
A matrix is a two-dimensional data structure.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
M <- matrix(0L, nrow=5,ncol=5) # 5x5 matrix of zeros
M
```

# Matrices
We can create a matrix by combining vectors using `cbind` or `rbind`.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
M1 <- cbind(v1,v2)
print(M1)
M2 <- rbind(v1, v2)
print(M2)
```

# Matrices
We can get particular values using two-dimensional indexing.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
i = 1 # row index
j = 1 # column index
M1[i,j] # Returns element 
M1[i,] # Returns row i
M1[,i] # Returns column i
```

# Data frames
Like its component vectors, a matrix contains data of the same type. If we have a mix of data types we generally want to use a ``data.frame``.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
df <- iris
df$a <- 1

df.2 <- df
df.2$b <- 2

df <- "wrggrwrg"
print(df)
```

# Data frames
Like its component vectors, a matrix contains data of the same type. If we have a mix of data types we generally want to use a ``data.frame``.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
colnames(iris) # gets column names, rownames will print index of each row
nrow(iris) # count rows
ncol(iris) # count columns
dim(iris) # count rows and columns
```


# Data frames
We can use indexing in the same way as lists to extract elements.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
iris$Sepal.Length[1] # Explicitly call column name
iris[[1]][1] # reference column using index
```

# Tibbles
A `tibble` is the `tidyverse` take on a data.frame. It is more "opinionated," which helps to maintain the integrity of your data. It also has some other updated features.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
library(nycflights13)
head(flights, n=5)
```

# Tibbles
We can easily convert any `data.frame` into a `tibble` and vice versa.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
library(tidyverse) # the library is required to use the as_tibble function
iris.t <- as_tibble(iris) # convert to tibble
class(iris.t)
iris.df <- as.data.frame(iris.t) # convert back to data.frame
class(iris.df)
```

# Style
## A note on style
- Not only do programming languages require a specific syntax to function, but there are also stylistic conventions
- There are packages you can use to automatically style your code (`styler` and `lintr`)
- See https://style.tidyverse.org/ for more info on R style

# Style
## Some style tips
- Naming
  - Use informative variable names
  - Keep names short
  - Maintain a consistent naming convention
- Use appropriate spacing to make code readable
  - e.g. `a = 1` is preferable to `a=1`
- Try to avoid extremely long expressions
  - Make complex functions modular (more next lecture)
  - Tidyverse uses the `%>%` operator to help with this (more next lecture)
- Comment your code for your future self and others
  
# Questions?


